Conversion of FFCS ICE fleet to EV: a data-driven approach

### 2020-11-09 Chapter10/rebuttal_letter.tex

\documentclass{paper}

%\setlength{\parskip}{0.11cm plus1mm minus1mm}
\setlength{\paperheight}{11in}
\setlength{\paperwidth}{8.5in}
\usepackage[top=0.70in, bottom=0.70in, left=0.75in, right=0.75in]{geometry}
\usepackage{soul}
\newcommand{\rev}[1]{\textbf{\textit{#1}}}
\newcommand{\ans}[1]{{\color{blue} #1}}
\newcommand{\tbd}[1]{\ul{#1}} % TO REMOVE AS LONG AS WE ADDRESS THE ISSUES IN PAPER

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[official]{eurosym}

%\usepackage{caption}

\usepackage[hyphenbreaks]{breakurl}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
%\usepackage{graphicx}
%\usepackage{textcomp}
%\usepackage{hyperref}
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}

\usepackage{xcolor}

%encoding
%--------------------------------------
%\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}
%-------------------------------------

\begin{document}

\noindent
Dear IEEE ISC2 2020 Chairs and Reviewers, \\

\noindent
This is the revision letter for the manuscript 1570661048, entitled ``On Scalability of Electric Car Sharing in Smart Cities''. \\

\noindent
First of all, let us thank the reviewers for their insightful comments. We did our best to address all the issues pointed out during the review round. In the following we report the point-to-point responses to the reviewers. 
%summarize the major changes introduced in this version:
%\begin{itemize}
%  \item We have clarified what is the influence of the operational area size on the customers usage;
%  \item We have proof-read the text to improve overall paper quality.

%\end{itemize}

\noindent
%We report at the end of this letter the references cited in the responses to the reviewer, keeping the same numbering as in the new manuscript. Please, see also the attached file (``diff.pdf'') which shows new content and modifications highlighted in blue.  \\

\begin{description}

 \item[\textbf{Response to Reviewer 1}] \hfill
\begin{description}
 \item[Comment 1] \rev{Is a unique global scaling factor enough for all the zones and all time periods? The role of this parameter should be explained better.}\\

% \ans{Per il nostro obbiettivo è sufficiente in quanto valutiamo il crescere della domanda supponendo una domanda stazionaria. Si potrebbe fare spazio variante/tempo variante, ma senza delle evidenze su quali zone/bin temporali sarebbe arbitrario. }\\

\ans{We agree with the reviewer that study the scalability by changing the spacial distribution of the trips other than the demand intensity would be another interesting research direction. We do no have enough data to assume different scale factors for different zones, i.e., an higher demand increase in some zones with respect to others.
%, we decided to keep the spatio-temporal patterns as the one observed in the input data. 
As such, in this work we are only studying how the system handles a growing amount of users' mobility requests while keeping the spatio-temporal patterns unchanged by employing a unique global scaling factor.
We clarified this aspect in the paper.
 %and comment this possibility in our future directions. 
  %as we mean to change only the average intensity of trips. Additional data would be thus needed to avoid arbitrary assumptions about temporal and spatial variations of mobility demand.
 }\\

 \item[Comment 2] \rev{Also the size of the variables is not clear. Why O-D matrices are 4D?} \\

\ans{The area of the city is divided into a two dimensional matrix composed of squared zones. At each zone we assign a couple of coordinates $(x,y)$. Hence, since each trip $i$ departs from a certain zone (origin $O_i$, described by two coordinates) and arrives to another zone (destination $D_i$, described by two coordinates), it is therefore characterized by two couple of coordinates, that can be represented as 4 scalars.

Thus, in order to model the OD matrix in each temporal slot, we fit a 4-dimensional KDE based on the aforesaid coordinates. We clarify this in the text.} \\

%DG: Attenzione che la correlazione OD non centra. Per tenerla dovremmo fare matrici separate - una per le partenze ed una contenente tutte le possibili destinazioni per partenza. In questo modo avremmo la correlazione tra partenze ed arrivi. Come facciamo noi semplicemente teniamo in considerazione del peso di ogni zona ma non delle correlazioni OD. E.g., Se avete 2 zone di origine (A,B) e due di destinazione (C,D). Dove A va sempre a C e B va sempre a D. Per come generiamo noi la matrice, potremmo avere dei viaggi che partono da A ed arrivano a D.

%Thus, in order to not lose the correlation of a certain origin with a certain destination at a given temporal slot, we fit a 4-dimensional KDE for each temporal slot describing origin-destination probabilities.

\item[Comment 3] \rev{Finally, does a 279-zone representation provide statistically significant results for a fleet of 400 vehicles? It is clear that it depends on the spatial distribution of the 180k trips, provided that it seems to be adequate compared to the distribution of the rental distance.}\\

%\ans{Se si riferisce che 180k trip siano sufficienti per far convergere le metriche (satisfied trips etc), sono sufficienti per 279 zone/400veicoli? Dovremmo controllare ma dovrebbe essere.
%\\
%Nella seconda parte della domanda sembra rispondersi da solo sul fatto che il numero di zone (e la zonizzazione) sia adeguata.
%-scelta 500m (survey sulla camminata)
%-rapporto viaggi sotto i 500m rispetto al totale
%-convergenza metriche controllata
%}\\

\ans{The number of zones in the city is a direct consequence of choosing 500m-sided squared zones. This choice is driven by a survey [a] in which it emerged that users are willing to walk about 500\,m at most in order to pick a vehicle from a shared fleet. Moreover, we observe from the data that almost all trips are not within the same zone. 
While the number of vehicles (400) is the one observed in original data, thus the one used by the operator during the considered time period.

We observed that in these conditions a 3-months simulation time (corresponding to 180k trips) is sufficient to let metrics converge to steady-state values.}\\

\item[Comment 4] \rev{Methodology of the demand model refers to a paper that is not yet published so that the reader cannot get full understanding of the methodology. Some assumptions should be discussed further.}\\

\ans{Given the space constraints we cannot add further details about the methodology and we refer the reader to the cited paper. To let other researchers read it before it will be fully available on IEEE Xplore, we share it upon request at the following link \url{https://www.researchgate.net/publication/344172879_Impact_of_Charging_Infrastructure_and_Policies_on_Electric_Car_Sharing_Systems}.}

\end{description}

\pagebreak

\item[\textbf{Response to Reviewer 2}] \hfill
\begin{description}

\item[Comment 1] \rev{The contributions of this paper should be stated in the Introduction section.}\\

\ans{We agree with the reviewer and we make the contributions clear in the Introductions.%We summarized the answers to the research questions making use of the demand model to project future or different scenarios and the cost-revenue model to evaluate the profitability of each configuration.
} \\

\item[Comment 2] \rev{This paper assumes that the electric cars do not pay for parking on street and for accessing limited traffic areas. This assumption is not very realistic. If this assumption is needed, this should be justified.}\\

\ans{
%Abbiamo qualche news o simile per giustificare questa assunzione? Altrimenti possiamo indicare che è una scelta fatta per incentivare l'utilizzo dei mezzi sharing elettrici da parte della municipalità per favorire l'economia green? Cambierebbe solo l'impatto dei veicoli andando ad aumentare a sfavore l'aumento del numero di auto.

According to [7] the Italian cities of Turin, Milan, Rome and Florence in 2017 did not required any fees to the clients for parking and accessing limited traffic areas  in case of electric vehicles.
Instead, internal combustion vehicles pay a cost for these two aspects which can vary from 550\,\euro{} to 1200\,\euro{} per vehicle and per year, depending on the city.
According to [b], in Madrid electric shared cars can park everywhere free of charge.

We are not aware of further agreements that the cities might have with the car sharing operators, hence it is hard to accurately estimate these figures. If we relax these hypotheses,  we should add a fee only to the vehicle cost in our cost-revenue model. Then, the trade off between number of charging poles and number of cars will favour even more the increase of number of poles.  %Anyway we estimate this cost to be much lower than the termic vehicle fee.
}\\

\item[Comment 3] \rev{The paper uses simulation to derive the insights. However, more details on the simulation implementation are needed. For example, what is the simulation platform?}\\

\ans{We used a custom simulator written in Python and based on SimPy, a discrete-event simulation library whose documentation is available at: https://simpy.readthedocs.io/en/latest/contents.html. Python version must be greater or equal to 3.6 in order to use SimPy library. We made these details explicit in the text. }\\

\item[Comment 4] \rev{Fig. 5, demand factor $->$ values of demand factor}\\

\ans{We fixed the sentence. }

\end{description}

\pagebreak

\item[\textbf{Response to Reviewer 3}] \hfill
\begin{description}

\item[Comment 1] \rev{The Authors illustrate some results obtained through a simulation performed starting from a realistic car sharing demand relevant to the city of Turin. The paper is interesting.
The Authors must add the Keywords that are missing
}\\

\ans{Thanks for the nice feedback. We added the missing keywords.}\\

\end{description}

\pagebreak

\item[\textbf{Response to Reviewer 4}] \hfill
\begin{description}

\item[Comment 1] \rev{I recommend the authors present the assumptions of the simulation which can help readers to understand the simulation model better.}\\

\ans{We made better explicit the assumptions in Section IV.a.}\\

\end{description}

\end{description}

\section*{References\footnote{Reference numbering is consistent with the article except for reference [a,b], which have been provided to the reviewers as additional documentation but not included in the article.}}

\begin{description}

\item[{[~a~]}] Herrmann, Sascha, Frederik Schulte, and Stefan Vo{\ss}. ``Increasing acceptance of free-floating car sharing systems using smart relocation strategies: a survey based study of car2go Hamburg".International conference on computational logistics. Springer, Cham, 2014.

\item[{[~b~]}] Lagadic, Marion, Alia Verloes, and Nicolas Louvet. "Can carsharing services be profitable? A critical review of established and developing business models." Transport Policy 77.C (2019): 68-78.

\item[{[~7~]}] M. Rossi, “Mobilita alternativa crack sharing,” ` Quattroruote, vol. 746,
2017. [Online]. Available: \url{http://shoped.it/shop/prodotto/quattroruoteottobre-2017/}

%@article{lagadic2019,
% author = {Lagadic, Marion and Verloes, Alia and Louvet, Nicolas},
% year = {2019},
% month = {02},
% pages = {},
% title = {Can carsharing services be profitable? A critical review of established and developing business models},
% volume = {77},
% journal = {Transport Policy},
% doi = {10.1016/j.tranpol.2019.02.006}
% }

% @article{Quattroruote,
% title = {Mobilità alternativa Crack sharing},
% author = {Mario Rossi},
% journal={Quattroruote},
% year={2017},
% volume={746},
% publisher={EditorialeDomus},
% url= {http://shoped.it/shop/prodotto/quattroruote-ottobre-2017/}
% }

\end{description}

\end{document}

### 2020-11-12 Chapter6/6_1_intro.tex
chapt

\section{Introduction}
\label{sec:intro}

@@

After quickly discussing related work in section~\ref{sec:related}, I present and characterize data in section~\ref{sec:data}, and the simulation model and tool in section~\ref{sec:Modelling}. section~\ref{sec:freefloating} discusses the impact of charging stations placement policies, while section~\ref{sec:results} compares return policies. section~\ref{sec:resPoles} presents the impact of concentrating or spreading charging stations in the city. Finally section~\ref{sec:conclusion} concludes the paper.

>>>>

\section{Introduction}
\label{sec:6_1_intro}

@@

After quickly discussing related work in section~\ref{sec:6_2_related}, I present and characterize data in section~\ref{sec:6_3_data}, and the simulation model and tool in section~\ref{sec:6_4_Modelling}. section~\ref{sec:6_6_freefloating} discusses the impact of charging stations placement policies, while section~\ref{sec:6_7_results} compares return policies. Section~\ref{sec:6_8_resPoles} presents the impact of concentrating or spreading charging stations in the city. Finally section~\ref{sec:6_9_conclusion} concludes the chapter.

### 2020-11-12 Chapter6/6_3_data.tex
chapt

\section{Data collection and characterization}
\label{sec:6_2_data}

@@

%\subsection{Data collection and filtering}
%\label{par:coll_and_filter}
%The modern FFCS provider like Car2go, DriveNow or GoGet use information systems to expose the position of the available cars through web services. Customers access these to check  which cars are available for a rental by using a smartphone app. For instance, Car2go offers public Application Programming Interface (API) \footnote{car2goAPI, \url{https://www.car2go.com/api/tou.htm} service subject to approval by car2Go. Approval granted in Sept. 2016 and discontinued in Jan. 2018.} to access the information about the system status. We use these APIs to collect data about actual rentals occurred over time following a streaming approach. First, we get the operative area for each city, i.e., the zone where cars can be rented and returned. The operative areas remain constant for several months.
%Next, we collect a snapshot every minute reporting the positions of all cars available for a rental, for each city. From each snapshot we save the car plates (used as car identifier) and its geographic coordinates.
%
%We developed UMAP~\cite{UMAP}, a software able to process these snapshots and rebuild, for each car, the history in terms of \emph{bookings} and \emph{parkings}.\footnote{UMAP can be downloaded from \url{https://github.com/MobilityPolito/UMAP}}
%A booking is an event describing a possible car rental $i$, characterized by the starting and final position $o(i)$ and $d(i)$, and starting and ending time $t_s(i)$ and $t_e(i)$. A parking describes a period of time when the car was parked and available for rentals; it is characterized by the location and the starting and final time. In a nutshell, we process the snapshots of the available cars and derive booking periods by observing which cars are available at each snapshot.
%Given a car, when it ``disappears'' from a snapshot, it means that someone has booked it -- so that the car is not available for other customers. A new booking is started. When the same car ``reappears'' back later on, it means someone has returned it. The booking is then ended.
%
%Not all bookings correspond to actual rentals:
%(i) a customer can book a car, and cancel the booking later on;
%(ii) the data collection may suffer from outages, so that some snapshots may miss some available cars;
%(iii) cars may go in maintenance, so that they disappear and never come back (or return after a long time).
%We developed data cleaning and filtering rules to extract actual \textit{rentals} from bookings. A booking is considered a valid rental if:
%(i) it lasts at least 3 minutes;
%(ii) the ending position is at least 700\,m far from the staring position, with both positions inside the city operative area;\footnote{This to account for possible errors in the GPS fixing, and to remove rentals started and ended in different cities.};
%(iii) its duration is smaller than 1\,h. These thresholds have been selected by domain knowledge of the data -- see~\cite{UMAP} for more details.
%Bookings that do not correspond to rentals are then merged with parkings events (since the car did not move).
%
%We started to collected data in December 2016 in all the 22 cities in which Car2go is present, and stopped our collection in February 2018. Overall, we obtained more than 28 millions rentals and parkings that we leverage for our studies.

\subsection{Temporal characterization}
I collected the data with the software described in chapter \ref{chap:2_dataset}. Reminding that not all bookings correspond to an actual rental because:
\begin{itemize}
	\item a customer can book a car, and cancel the booking later on;
	\item the data collection may suffer from outages, so that some snapshots may miss some available cars;
	\item cars may go in maintenance, so that they disappear and never come back (or return after a long time)
\end{itemize}
%(i) a customer can book a car, and cancel the booking later on;
%(ii) the data collection may suffer from outages, so that some snapshots may miss some available cars;
%(iii) cars may go in maintenance, so that they disappear and never come back (or return after a long time).
I developed data cleaning and filtering rules to extract actual \textit{rentals} from bookings. A booking is considered a valid rental if:
\begin{itemize}
	\item it lasts at least 3 minutes;
	\item the ending position is at least 700\,m far from the staring position, with both positions inside the city operative area;\footnote{This to account for possible errors in the GPS fixing, and to remove rentals started and ended in different cities.};
	\item its duration is smaller than 1\,h. These thresholds have been selected by domain knowledge of the data -- see~\cite{ciociolaumap} for more details.
	\item \mc{rivedere capitolo appropriato con vari link, e linkare anche le def}
\end{itemize}
Bookings that do not correspond to rentals are then merged with parkings events (since the car did not move).

%I started to collected data in December 2016 in all the 22 cities in which Car2go is present, and stopped our collection in February 2018. Overall, we obtained more than 28 millions rentals and parkings that we leverage for our studies.

@@

\begin{figure}[t]
    \begin{center}
        \begin{subfigure}{\textwidth}
            \includegraphics[width=\columnwidth]{figures/bookings_per_day.pdf}
            \caption{Number of rentals per day.}
            \label{fig:rentals_per_day}
        \end{subfigure}
         \begin{subfigure}{\textwidth}
             \includegraphics[width=\columnwidth]{figures/aggBookginfsPerHour.pdf}
             \caption{Average number of rentals per hour.}
             \label{fig:rentals_per_hour}
         \end{subfigure}
         \caption{Rental trends in different cities.}
         \label{fig:rentals_trends}
\end{center}
\end{figure}

\begin{figure}[t]
%     \begin{center}
            \begin{subfigure}{\textwidth}
            \includegraphics[width=\columnwidth]{figures/CDF_aggregate_RentalsDuration.pdf}
           % \caption{CDF of driving duration}
            %\label{fig:cdf_driving_duration}
\vspace{-0.3cm}
        \end{subfigure}\\
        \begin{subfigure}{\textwidth}
             \includegraphics[width=\columnwidth]{figures/CDF_aggregate_RentalsDistance.pdf}
           %  \caption{CDF of driving distance}
            % \label{fig:cdf_driving_distance}
\vspace{-0.3cm}
         \end{subfigure}\\
         \begin{subfigure}{\textwidth}
             \includegraphics[width=\columnwidth]{figures/CDF_aggregate_ParkingsDuration.pdf}
           %  \caption{CDF of parking duration}
            % \label{fig:cdf_parking_duration}
         \end{subfigure}

         \caption{Car Sharing usage habit characterization.}
         \label{fig:cdf_characterization}
% \end{center}
\end{figure}

To give the intuition of the system, I first provide a characterization of actual usage patterns by current FFCS customers in each city. I focus first on the temporal characteristics.
Figure~\ref{fig:rentals_trends} reports the rental trend. More in details, figure~\ref{fig:rentals_per_day} shows the number of recorded rentals for each day in the considered period. Usage similarity is striking, with Milan, Berlin and Vancouver that have more rentals per day than Turin. This intense usage, justify the difference in fleet size between the cities, with these three having at least twice as much cars with respect to Turin (see Tab.~\ref{tab:summary}). These first results highlights the importance of extending the users' pattern analyses of same provider on different cities. A second interesting aspect is the presence of a weekly pattern: in correspondence of the weekends the number of rentals drops of about 30\%. This is justified by the fact that during the working days cars are used for commuting. Moreover, non-stationary events due to holidays or strikes are visible, e.g., October 6th in Milan due to a public transport strike.\footnote{The sudden fall around the October $2^{th}$ is due to system outage that caused a lack of data.}

\begin{figure*}[t!]
    \begin{center}
        \begin{subfigure}{\textwidth}
		    \begin{center}
            \includegraphics[width=0.495\columnwidth]{figures/BerlinoAvgParking.pdf}
            \includegraphics[width=0.495\columnwidth]{figures/BerlinoMaxParking.pdf}                
            \captionsetup{justification=centering}
            \caption{Berlin: average parking time (left) and total number of  parkings (right).}
            \label{fig:heatmap_Berlin}
            \end{center}
        \end{subfigure}
        \begin{subfigure}{\textwidth}
		    \begin{center}
            %\vspace{20pt}
            \includegraphics[width=0.495\columnwidth]{figures/VancouverAvgParking.pdf}
            \includegraphics[width=0.495\columnwidth]{figures/VancouverMaxParking.pdf}
            \captionsetup{justification=centering}
            \caption{Vancouver: average parking time (left) and total number of parkings (right).}
            \label{fig:heatmap_vancouver}
            \end{center}
        \end{subfigure}
          \caption{Heat map of average parking time and total number of parking. The warmer the color, the higher the value.}
         \label{fig:heatmaps}
\end{center}
\end{figure*}

To deeply analyze customers' habits, I detail the average number of rentals per hour in figure~\ref{fig:rentals_per_hour}, separately per working-days (WD, solid line) and per weekends (WE, dashed line). Each curve reports the number of rentals for each hour, considering the average number over the same hour in the dataset.
Firstly, notice the usage peaks during commuting times. These happen at different times for different cities, e.g., 8am for Turin, Milan and Berlin vs 7am for Vancouver, following local commuting habits. Secondly, notice how the evening and night usage tend to be larger during  weekends than working days. This reflects the different usage patterns at night, when cars are used to reach areas dense of pubs and nightlife.
At last, observe again the different patterns in different cities. For instance, the average number of rentals in Vancouver and Berlin during weekend mornings is higher than during working days. This does not happen in Italian cities. The charging station placement design must thus weight this different needs and non-stationary patterns.

Given the goals of deriving guidelines for charging station placement policies, I focus now on the characterization of three important metrics:
(i) for how long customers rent a car,
(ii) how far they drive,
(iii) how long cars usually stay parked.
The former two metrics guide the battery discharging properties, while the latter metric is fundamental to understand battery charging opportunities. Given the data-set does not have any information on car position during a rental, I compute the travel distance by assuming the customer went directly from the origin to the destination. This is indeed compatible with the duration of rentals (see below). I use Google Map service to compute a correcting factor to be applied on the euclidean distance, \mc{citare sezione apposita}.

Figure~\ref{fig:cdf_characterization} reports the Cumulative Distribution Function (CDF) of the rentals duration (top), driving distance (middle) and parking duration (bottom).
The size of the city has a clear impact, with Turin that has the shortest trips, and Berlin the longest. 
Rental duration is in general very short, leading to the intuition that drivers tend to minimize the rental time (and cost). Traveled distance is fundamental to understand the battery consumption: The maximum driving distance sets the minimum battery charge to sustain that trip. Looking at the middle plot in digure~\ref{fig:cdf_characterization}, I observe that in Berlin the longest trips are twice as long as the longest trips in other cities. Therefore, the same battery constraints would not fit for all cities.

@@

At last, bottom plot of figure~\ref{fig:cdf_characterization} details the duration of the parking periods. Interestingly, 50\% of parkings lasts less than 22 minutes in Berlin, testifying a very high system utilization. In Turin, the median grows to 42 minutes, still showing that most cars are parked for short time. Yet, the long tail of the CDF (note the log scale on x-axis) suggests that there is a sizeable fraction of parkings that last for 5 or more hours. Cars parked in the periphery typically at night, where the demand is lower, belongs to this fraction.
\begin{table}[h!]
\begin{center}
\scriptsize
\caption{Main characteristics of data.\label{tab:summary}}
\setlength\tabcolsep{5pt} % default value: 6pt reduce cell padding
\begin{tabular}{|l|l|l|l|c|c|c|c|c|l|}
\hline
\!\multirow{3}{*}{City} &\!\multirow{3}{*}{Rental}\!& \multirow{2}{*}{Fleet} &  \multicolumn{2}{|c|}{Rental}  &  \multicolumn{2}{|c|}{Rental}  &  \multicolumn{2}{|c|}{Parking}  &\!\multirow{3}{*}{Zones}\!\\ 
 &  & \multirow{2}{*}{Size}  & \multicolumn{2}{|c|}{Time~[min]}  & \multicolumn{2}{|c|}{Dist.~[km]} & \multicolumn{2}{|c|}{Time~[h]} &    \\ \cline{4-9} %&  \multirow{2}{*}{Size[GB]}
 & &  &\!Avg\!&\!Med\!&\!Avg\!&\!Med\!&\!Avg\!&\!Med\!& \\ \hline
\hline
       \!Torino       & 125k & 377 & 21 & 20 & 3.96 & 3.36 & 3:17 & 0:42 & 261 \\ \hline
\!Milano       & 320k & 739 & 25 & 24 & 4.15 & 3.66 & 2:21 & 0:24 & 549 \\ \hline
\!Berlino      & 342k & 900 & 29 & 28 & 6.22 & 5.24 & 2:23 & 0:22 & 833 \\ \hline
\!Vancouver\!  & 317k & 941 & 26 & 24 & 4.70 & 3.98 & 2:54 & 0:22 & 532 \\ \hline

@@

Figure~\ref{fig:heatmaps} shows the above metrics for Berlin and Vancouver using a heat map -- with blue and red corresponding to the minimum and maximum values, respectively. Focus on Berlin first - figure~\ref{fig:heatmap_Berlin}. Left plot shows the average parking time. Results clearly show that cars stay parked for very short time in busy downtown areas, while cars stay parked for longer time in the periphery (corresponding to the head and tail of the CDF in figure~\ref{fig:cdf_characterization}, respectively).

@@

In a nutshell, in busy areas, the average parking time is short, and number of rentals and parkings recorded is high. This reflects the specific usage of FFCS according to which cars move to downtown areas in the morning, then are rented to move within central areas, and finally are driven back to the periphery at the end of the day.
Similar results apply to all cities -- see Fig.~\ref{fig:heatmap_vancouver} which details Vancouver statistics.

>>>>

\section{Data collection and characterization}
\label{sec:6_3_data}

@@

\subsection{Temporal characterization}
I collected the data with the software described in chapter \ref{chap:2_dataset}. Recalling that the basic dataset entity is the \textit{bookings}, formalized in definition \ref{def:booking} . Not all bookings correspond to an actual real user's trip. Indeed it is possible to observe the booking dataset is composed by \textit{reservations} and \textit{rentals}, defined respectively in definition \ref{def:reservation} and definition \ref{def:rental}. This ambiguity is due to the high degree of freedom of a FFCS system:
\begin{itemize}
	\item a customer can book a car, and cancel the booking later on;
	\item the data collection may suffer from outages, so that some snapshots may miss some available cars;
	\item cars may go in maintenance, so that they disappear and never come back (or return after a long time)
\end{itemize}

To tackle this problem, I developed data cleaning and filtering rules to extract actual \textit{rentals} from bookings date-set. A rental is identified when:
\begin{itemize}
	\item it lasts at least 3 minutes;
	\item the ending position is at least 700\,m far from the staring position, with both positions inside the city operative area;\footnote{This to account for possible errors in the GPS fixing, and to remove rentals started and ended in different cities.};
	\item its duration is smaller than 1\,h. These thresholds have been selected by domain knowledge of the data -- see~\ref{chap:3_charact} for more details.

\end{itemize}
Bookings that do not correspond to rentals are then merged with parkings events (since the car did not move).

@@

\begin{figure}[t]
    \begin{center}
        \begin{subfigure}{\textwidth}
            \includegraphics[width=\columnwidth]{figures/bookings_per_day.pdf}
            \caption{Number of rentals per day.}
            \label{fig:6_3_rentals_per_day}
        \end{subfigure}
         \begin{subfigure}{\textwidth}
             \includegraphics[width=\columnwidth]{figures/aggBookginfsPerHour.pdf}
             \caption{Average number of rentals per hour.}
             \label{fig:6_3_rentals_per_hour}
         \end{subfigure}
         \caption{Rental trends in different cities.}
         \label{fig:6_3_rentals_trends}
\end{center}
\end{figure}

\begin{figure}[t]
%     \begin{center}
            \begin{subfigure}{\textwidth}
            \includegraphics[width=\columnwidth]{figures/CDF_aggregate_RentalsDuration.pdf}
           % \caption{CDF of driving duration}
            %\label{fig:6_3_cdf_driving_duration}
\vspace{-0.3cm}
        \end{subfigure}\\
        \begin{subfigure}{\textwidth}
             \includegraphics[width=\columnwidth]{figures/CDF_aggregate_RentalsDistance.pdf}
           %  \caption{CDF of driving distance}
            % \label{fig:6_3_cdf_driving_distance}
\vspace{-0.3cm}
         \end{subfigure}\\
         \begin{subfigure}{\textwidth}
             \includegraphics[width=\columnwidth]{figures/CDF_aggregate_ParkingsDuration.pdf}
           %  \caption{CDF of parking duration}
            % \label{fig:6_3_cdf_parking_duration}
         \end{subfigure}

         \caption{Car Sharing usage habit characterization.}
         \label{fig:6_3_cdf_characterization}
% \end{center}
\end{figure}

To give the intuition of the system, I first provide a characterization of actual usage patterns by current FFCS customers in each city. I focus first on the temporal characteristics.
Figure~\ref{fig:6_3_rentals_trends} reports the rental trend. More in details, figure~\ref{fig:6_3_rentals_per_day} shows the number of recorded rentals for each day in the considered period. Usage similarity is striking, with Milan, Berlin and Vancouver that have more rentals per day than Turin. This intense usage, justify the difference in fleet size between the cities, with these three having at least twice as much cars with respect to Turin (see table.~\ref{tab:6_3_summary}). These first results highlights the importance of extending the users' pattern analyses of same provider on different cities. A second interesting aspect is the presence of a weekly pattern: in correspondence of the weekends the number of rentals drops of about 30\%. This is justified by the fact that during the working days cars are used for commuting. Moreover, non-stationary events due to holidays or strikes are visible, e.g., October 6th in Milan due to a public transport strike.\footnote{The sudden fall around the October $2^{th}$ is due to system outage that caused a lack of data.}

\begin{figure*}[t!]
    \begin{center}
        \begin{subfigure}{\textwidth}
		    \begin{center}
            \includegraphics[width=0.495\columnwidth]{figures/BerlinoAvgParking.pdf}
            \includegraphics[width=0.495\columnwidth]{figures/BerlinoMaxParking.pdf}                
            \captionsetup{justification=centering}
            \caption{Berlin: average parking time (left) and total number of  parkings (right).}
            \label{fig:6_3_heatmap_Berlin}
            \end{center}
        \end{subfigure}
        \begin{subfigure}{\textwidth}
		    \begin{center}
            %\vspace{20pt}
            \includegraphics[width=0.495\columnwidth]{figures/VancouverAvgParking.pdf}
            \includegraphics[width=0.495\columnwidth]{figures/VancouverMaxParking.pdf}
            \captionsetup{justification=centering}
            \caption{Vancouver: average parking time (left) and total number of parkings (right).}
            \label{fig:6_3_heatmap_vancouver}
            \end{center}
        \end{subfigure}
          \caption{Heat map of average parking time and total number of parking. The warmer the color, the higher the value.}
         \label{fig:6_3_heatmaps}
\end{center}
\end{figure*}

To deeply analyze customers' habits, I detail the average number of rentals per hour in figure~\ref{fig:6_3_rentals_per_hour}, separately per working-days (WD, solid line) and per weekends (WE, dashed line). Each curve reports the number of rentals for each hour, considering the average number over the same hour in the dataset.
Firstly, notice the usage peaks during commuting times. These happen at different times for different cities, e.g., 8am for Turin, Milan and Berlin vs 7am for Vancouver, following local commuting habits. Secondly, notice how the evening and night usage tend to be larger during  weekends than working days. This reflects the different usage patterns at night, when cars are used to reach areas dense of pubs and nightlife.
At last, observe again the different patterns in different cities. For instance, the average number of rentals in Vancouver and Berlin during weekend mornings is higher than during working days. This does not happen in Italian cities. The charging station placement design must thus weight this different needs and non-stationary patterns.

Given the goals of deriving guidelines for charging station placement policies, I focus now on the characterization of three important metrics:
(i) for how long customers rent a car,
(ii) how far they drive,
(iii) how long cars usually stay parked.
The former two metrics guide the battery discharging properties, while the latter metric is fundamental to understand battery charging opportunities. Given the data-set does not have any information on car position during a rental, I compute the travel distance by assuming the customer went directly from the origin to the destination. This is indeed compatible with the duration of rentals (see below). I used Google Map service to compute a correcting factor to be applied on the euclidean distance as described in chapter \ref{chap:3_charact}.

Figure~\ref{fig:6_3_cdf_characterization} reports the Cumulative Distribution Function (CDF) of the rentals duration (top), driving distance (middle) and parking duration (bottom).
The size of the city has a clear impact, with Turin that has the shortest trips, and Berlin the longest. 
Rental duration is in general very short, leading to the intuition that drivers tend to minimize the rental time (and cost). Traveled distance is fundamental to understand the battery consumption: The maximum driving distance sets the minimum battery charge to sustain that trip. Looking at the middle plot in digure~\ref{fig:6_3_cdf_characterization}, I observe that in Berlin the longest trips are twice as long as the longest trips in other cities. Therefore, the same battery constraints would not fit for all cities.

@@

At last, bottom plot of figure~\ref{fig:6_3_cdf_characterization} details the duration of the parking periods. Interestingly, 50\% of parkings lasts less than 22 minutes in Berlin, testifying a very high system utilization. In Turin, the median grows to 42 minutes, still showing that most cars are parked for short time. Yet, the long tail of the CDF (note the log scale on x-axis) suggests that there is a sizeable fraction of parkings that last for 5 or more hours. Cars parked in the periphery typically at night, where the demand is lower, belongs to this fraction.
\begin{table}[h!]
\begin{center}
\scriptsize
\caption{Main characteristics of data.\label{tab:6_3_summary}}
\setlength\tabcolsep{5pt} % default value: 6pt reduce cell padding
\begin{tabular}{|l|l|l|l|c|c|c|c|c|l|}
\hline
\!\multirow{3}{*}{City} &\!\multirow{3}{*}{Rental}\!& \multirow{2}{*}{Fleet} &  \multicolumn{2}{|c|}{Rental}  &  \multicolumn{2}{|c|}{Rental}  &  \multicolumn{2}{|c|}{Parking}  &\!\multirow{3}{*}{Zones}\!\\ 
 &  & \multirow{2}{*}{Size}  & \multicolumn{2}{|c|}{Time~[min]}  & \multicolumn{2}{|c|}{Dist.~[km]} & \multicolumn{2}{|c|}{Time~[h]} &    \\ \cline{4-9} %&  \multirow{2}{*}{Size[GB]}
 & &  &\!Avg\!&\!Med\!&\!Avg\!&\!Med\!&\!Avg\!&\!Med\!& \\ \hline
\hline
       \!Torino       & 125k & 377 & 21 & 20 & 3.96 & 3.36 & 3:17 & 0:42 & 261 \\ \hline
\!Milano       & 320k & 739 & 25 & 24 & 4.15 & 3.66 & 2:21 & 0:24 & 549 \\ \hline
\!Berlino      & 342k & 900 & 29 & 28 & 6.22 & 5.24 & 2:23 & 0:22 & 833 \\ \hline
\!Vancouver\!  & 317k & 941 & 26 & 24 & 4.70 & 3.98 & 2:54 & 0:22 & 532 \\ \hline

@@

Figure~\ref{fig:6_3_heatmaps} shows the above metrics for Berlin and Vancouver using a heat map -- with blue and red corresponding to the minimum and maximum values, respectively. Focus on Berlin first - figure~\ref{fig:6_3_heatmap_Berlin}. Left plot shows the average parking time. Results clearly show that cars stay parked for very short time in busy downtown areas, while cars stay parked for longer time in the periphery (corresponding to the head and tail of the CDF in figure~\ref{fig:6_3_cdf_characterization}, respectively).

@@

In a nutshell, in busy areas, the average parking time is short, and number of rentals and parkings recorded is high. This reflects the specific usage of FFCS according to which cars move to downtown areas in the morning, then are rented to move within central areas, and finally are driven back to the periphery at the end of the day.
Similar results apply to all cities -- see Fig.~\ref{fig:6_3_heatmap_vancouver} which details Vancouver statistics.
